---
title: "PA1_Template"
author: "Jason Crain"
date: "January 21, 2016"
output: html_document
---

# Loading and preprocessing

Load up the activity data from the git repo. Take a look at the contents to
figure out what we need to do.

```{r echo = TRUE }
DataSet <- read.csv("activity.csv")
head(DataSet)
```

Looks like there's a lot of NA data. Take a quick look for non NA data.

```{r echo = TRUE}
head(DataSet[ which (is.na(DataSet$steps) == FALSE), ])

```

OK.. so lots of zeros, as well.. what else is there?

```{r echo = TRUE}
head(DataSet[ which (DataSet$steps > 5), ])

```

OK. Found good data. I don't think there's any need to preprocess the data.

# Daily mean and activity

To start looking into the daily mean and the activity across different intervals,
take a subset of the data, removing NA values.
```{r echo = TRUE}
NoNASet <- DataSet[ which (is.na(DataSet$steps) == FALSE), ]
```

After that, get the dates and interval values for use later, and make two frames.
One will contain the sum of steps per day, and the other will contain the sum of
steps per interval.
```{r echo = TRUE}
# Get dates and intervals from the set for use later
Dates <- as.Date(unique(NoNASet$date))
Intervals <- unique(NoNASet$interval)

# Setup two frames for use to show by date and by interval
StepsByDayFrame <- data.frame(stringsAsFactors = FALSE, DATE = Dates)
StepsByIntervalFrame <- data.frame(stringsAsFactors = FALSE, INTERVAL = Intervals)
```

Now, we're going to want to use a function so that we can get aggregate data.
After we make the functions, use sapply to apply them to each unique value for
date and interval.
```{r echo = TRUE}
# Function to get steps and lapply to get a list
GetStepsByDate <- function(n) {
    sum(NoNASet[ which (as.Date(NoNASet$date) == n), 1])
}
GetStepsByInterval <- function(i) {
    sum(NoNASet [ which (NoNASet$interval == i), 1])
}

stepsvec <- sapply(StepsByDayFrame$DATE, GetStepsByDate)
intervalvec <- sapply(StepsByIntervalFrame$INTERVAL, GetStepsByInterval)
```

Now, let's finish making the frames and check the data to see how we did.
```{r echo = TRUE}
# check the frames for errors
StepsByDayFrame <- cbind(StepsByDayFrame, datemean = stepsvec, stringsAsFactors = FALSE)
head(StepsByDayFrame)

StepsByIntervalFrame <- cbind(StepsByIntervalFrame, intervalmean = intervalvec, 
                              stringsAsFactors = FALSE)
head(StepsByIntervalFrame)
```
